---
title: "Loan Long and Prosper"
author: "Dylan Petrusma"
date: "`r Sys.Date()`"
output: html_document
---

Loan Long and Prosper by Dylan Petrusma
========================================================

# Abstract
This report explores a dataset containing borrower and loan data for
approximately 110,000 loans from Prosper, a Peer-to-Peer lending service.

# Dataset

### Preparations

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
#install.packages('GGally')
library(GGally)
library(zoo)
#This is for proper rounding
#install.packages("broman")
library(broman)
library(scales)
#This is for counting the days in a year
#install.packages("lubridate")
library(lubridate)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Do_not_forget}
#Here is a cool thing: `r myround(-0.01564787, 2)`
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data and convert the date/datetime columns from strings
prosper_loans <- read.csv(
    "C:\\Users\\Dylan\\Downloads\\Large Files\\prosperLoanData.csv", na.strings
    = c("", "NA"))

prosper_loans$ListingCategory <- factor(prosper_loans$ListingCategory..numeric.)

prosper_loans$LoanOriginationDate <- as.Date(prosper_loans$ListingCreationDate)
prosper_loans$FirstRecordedCreditLine <- as.Date(
  prosper_loans$FirstRecordedCreditLine)
prosper_loans$ListingCreationDate <- strptime(prosper_loans$ListingCreationDate,
                                             "%Y-%m-%d %H:%M:%S")
prosper_loans$ClosedDate <- strptime(prosper_loans$ListingCreationDate,
                                    "%Y-%m-%d %H:%M:%S")
prosper_loans$DateCreditPulled <- strptime(prosper_loans$DateCreditPulled,
                                          "%Y-%m-%d %H:%M:%S")

prosper_loans$IncomeRange <- factor( prosper_loans$IncomeRange, levels = c(
  "Not displayed", "Not employed", "$0", "$1-24,999", "$25,000-49,999",
  "$50,000-74,999", "$75,000-99,999", "$100,000+"))
#print(levels(prosper_loans$IncomeRange))

prosper_loans$CreditGrade <- factor( prosper_loans$CreditGrade, levels = c(
  "NA", "NC", "HR", "E", "D", "C", "B", "A", "AA"))
#print(levels(prosper_loans$CreditGrade))

prosper_loans$ProsperRating..Alpha. <-
  factor( prosper_loans$ProsperRating..Alpha., levels = c(
  "NA", "HR", "E", "D", "C", "B", "A", "AA"))
#print(levels(prosper_loans$ProsperRating..Alpha.))

prosper_loans$CreditScoreRangeLowerFactor <- factor(prosper_loans$CreditScoreRangeLower)

str(prosper_loans)
```

# Univariate Plots Section

The variables of most interest to me in the beginning were the loan amount, and
the loan term, and I started by checking the basic statistics of them, followed
by a few plots of other variables that seemed interesting.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_1}
# Some quick structure checks to get a feel for the loan amounts and terms.
#Simple counts
summary(prosper_loans$LoanOriginalAmount)
summary(prosper_loans$Term)

ggplot(data = prosper_loans, aes(x = LoanOriginalAmount)) +
  geom_histogram(binwidth = 1000)

ggplot(data = prosper_loans, aes(x = Term)) +
  geom_histogram(binwidth = 6) +
  scale_x_continuous(limits = c(0,66), breaks = seq(0,66,6))
```

Clearly the loan amounts are positively skewed and clustered around some round
numbers (roughly $5,000, $10,000, $15,000, $20,000 and $25,000 from looking at
the chart), but the terms only fall into only 3 ranges: 12, 36, and 60 months.

Next I wanted to look at the number of loans taken changing over time and the
current statuses of all of the loans

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_1a}
ggplot(data = prosper_loans, aes(x = as.yearmon(LoanOriginationDate))) +
  geom_area(stat = "count") +
  scale_x_yearmon()

ggplot(data = prosper_loans, aes(x = LoanStatus)) +
  geom_bar(stat = "count") +
  scale_y_log10() +
  coord_flip()
```

The post-2008-GFC drop in loan numbers is very clear to see, though from 2009
sometime the number has been going up and up. I wonder if certain ratios have
changed over time, such as the debt-to-income ratio of borrowers? Luckily for
Prosper, very few loans are in an overdue state.

Next I wanted to explore a few more categorical variables.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_1b}
ggplot(data = prosper_loans, aes(x = CreditGrade)) +
  geom_bar(stat = "count")

ggplot(data = subset(prosper_loans, !is.na(prosper_loans$CreditGrade)),
       aes(x = CreditGrade)) +
  geom_bar(stat = "count")

ggplot(data = subset(prosper_loans, !is.na(prosper_loans$BorrowerState)),
       aes(x = BorrowerState)) +
  geom_bar(stat = "count") +
  theme(axis.text.x = element_text(angle = -90, hjust = 0))

ggplot(data = prosper_loans, aes(x = ListingCategory..numeric.)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(0,20,1)) +
  scale_y_log10()

ggplot(data = subset(prosper_loans, !is.na(prosper_loans$MonthlyLoanPayment)),
       aes(x = MonthlyLoanPayment)) +
  geom_histogram() +
  scale_y_log10()
```

I was surprised to see such a distribution of credit grade. My assumption was
that the most common category would be AA, as the company would seek good
quality clients, but it actually follows more of a normal distribution, once I
took out the NA entries.

The resident state of the borrowers seems fairly randomly distributed, as I
expected, though I am a little curious as to why California has more than twice
the next most common state. Most likely it's simply a factor of population -
California is the most populous state in the USA.

The most common loan category is 1, which corresponds to "Debt Consolidation".
I chose a log 10 scale for the y-axis as the number of loans for most categories
is miniscule compared to categories 0, 1, 2, 3, and 7, so this lets me see a
little better.

The monthly loan payments are rather positively skewed, which matches my
intuition - the most common loans would be smaller, I would think, but not too
small or the lenders cannot make a profit. I chose a log 10 scale for the y-axis
as the number of loans with payments more than $1,000 is miniscule compared to
the sub-$500 monthly payment loans.


After looking at the distribution of the number of loans over time previously,
I was interested to see any seasonable patterns by dropping the year. However,
you can see the following first plot is not so helpful, being far too granular,
and so I dropped the day as well and only looked at the month.

```{r echo=FALSE, message=FALSE, warning=FALSE, OneMorePlot}
prosper_loans$LoanOriginationDateDaysFromJanFirst <- yday(prosper_loans$LoanOriginationDate) - 1

ggplot(data = prosper_loans, aes(x = LoanOriginationDateDaysFromJanFirst)) +
  geom_histogram(stat = "count") +
  scale_x_continuous(breaks = seq(0,365,30))

#This is way too busy, so I'll try a few other things
prosper_loans$LoanOriginationMonthName <- format(
  prosper_loans$LoanOriginationDate, format = "%b")
#prosper_loans$LoanOriginationMonth <- as.double(format(
#  prosper_loans$LoanOriginationDate, format = "%Y%m"))
prosper_loans$LoanOriginationMonth <- format(prosper_loans$LoanOriginationDate,
                                             format = "%Y%m")

ggplot(data = prosper_loans, aes(x = LoanOriginationMonthName)) +
  geom_histogram(stat = "count")

#One more version
ggplot(data = prosper_loans, aes(x = LoanOriginationMonthName)) +
  geom_line(stat = "count", aes(color = format(
    prosper_loans$LoanOriginationDate, "%Y"),
    group = format(prosper_loans$LoanOriginationDate, "%Y"))) +
  xlab("Loan Month") +
  ylab("Number of Loans") +
  scale_color_discrete(name = "Year")
```

When looking at the total count for all years (the second histogram just above)
I can see a slight variation throughout the year - low point in April, and high
point in January, with a fairly smooth gradiation between the two points in both
directions.

The second (line) chart shows that within each year there is quite a bit of
variation. In 2013 the number of loans very quickly picked up, though it seems
to have dropped quite sharply during the first quarter of 2014.

```{r echo=FALSE, message=FALSE, warning=FALSE, OneMorePlot_2}
ggplot(data = prosper_loans, aes(x = IncomeRange) ) +
  geom_bar(stat = "count") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0))
```

As I would expect, not many of the loans are for customers with no income, not
employed, or incomes below $25,000 - most of their loans fall in the average
range of $25,000 - $75,000 income levels.

# Univariate Analysis

### What is the structure of your dataset?

There are 113,937 loans in the dataset with 85 features, including the few I
created myself.

### What is/are the main feature(s) of interest in your dataset?

For me, the main features of interest are the trends over time of some of the
categorical features, such as the loan type and credit grade, as well as the
numerical features, such as debt-to-income ratio and loan amount.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I will also look at the Prosper Rating and Credit Grade to investigate the loan
amount, as well as the credit score to look for patterns in the categorical
features mentioned above.

### Did you create any new variables from existing variables in the dataset?

I did, but only a few. I wanted to bin the loan origination date a little
differently, so I created features of the loan date without the day, i.e., year
and month only, loan date without the day or year, i.e., number of days from Jan
1st of that year, and the loan month.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

No unusually shaped distributions, though a few were not quite what I expected,
having a peak at a different value for the credit grade, for example.

# Bivariate Plots Section

Next I wanted to understand a little better how the loan amounts varied across
the listing categories and employment statuses.

It's interesting but there doesn't seem to be much correlation here. As I would
expect, "Employed" appears to have the largest spread, as it would probably be
the most used category. What is interesting is to see the $25,000 loan is the
maximum for most employment statuses - perhaps that is the maximum Prosper will
lend you unless you are employed full time?

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_2}
# Loan amount and numbers by category and employement status
ggplot(data = prosper_loans, aes(x = ListingCategory, y = LoanOriginalAmount)) +
  geom_boxplot()

ggplot(data = prosper_loans, aes(x = EmploymentStatus,
                                 y = LoanOriginalAmount)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  coord_flip()
```

Looking at the loan categories, I can see that categories 4 and 5 have the
lowest median, which correspond to "Personal Loan" and "Student Use",
respectively. This is to be expected as a student is less likely to be able to
afford a large loan, and personal loans that do not fall into another category
would tend to be for smaller costs, such as electronics or whitegoods. The
category with the lowest maximum value is 10, which is "Cosmetic Procedure". I'm
not sure if that is due to the nature of the loans, or if it is simply a
consequence of a small number of loan in that category; the chart above shows
it to have the smallest number of outliers, in this case 1.

Category 1, i.e., "Debt Consolidation", shows the highest median, which is also
expected as by the nature of the loan you are combining several others into one.

Next I wanted to see if there was an obvious relationship between credit rating
and the borrower rate - I was expecting a strong one.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_1}
# I wanted to see if the expected relationship between credit rating and loan
# rate manifests. There are some entries with very low scores that I believe are
# errors, hence I filter out below 100
ggplot(data = subset(prosper_loans, CreditScoreRangeLower > 100),
       aes(y = CreditScoreRangeLower, x = BorrowerRate)) +
  #geom_ribbon()
  geom_point(alpha = 0.05) +
  geom_smooth()

ggplot(data = subset(prosper_loans, CreditScoreRangeUpper > 100),
       aes(y = CreditScoreRangeUpper, x = BorrowerRate)) +
  #geom_ribbon()
  geom_point(alpha = 0.05) +
  geom_smooth()

# ggplot() +
#   geom_smooth(data = subset(prosper_loans, CreditScoreRangeUpper > 100),
#               aes(y = CreditScoreRangeUpper, x = BorrowerRate,
#                   color = I("red"))) +
#   geom_smooth(data = subset(prosper_loans, CreditScoreRangeLower > 100),
#               aes(y = CreditScoreRangeLower, x = BorrowerRate,
#                   color = I("blue"))) +
#   ylab("Credit Range")
```

There seems to be a slight correlation apparent between borrower rate and lower
credit rating. Interestingly, there are customers with credit scores of 500 and
rates of 0.05. I wonder if they are for certain loan types or employment
statuses? This question will be explored in the Mutlivariate plots section.

Next I would like to look at the loan amounts as time passes, as well as the
debt to income ratio accepted

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_4}
ggplot(data = prosper_loans, aes(x = as.yearmon(LoanOriginationDate),
                                 y = LoanOriginalAmount)) +
  geom_line(stat = "summary", fun.y = mean) +
  scale_x_yearmon() +
  geom_smooth()

ggplot(data = prosper_loans, aes(x = as.yearmon(LoanOriginationDate),
                                 y = DebtToIncomeRatio)) +
  geom_line(stat = "summary", fun.y = mean) +
  scale_x_yearmon() +
  geom_smooth() +
  ylim(0, 0.3)
```

Both of these factors appear to be trending upwards as time goes by, though the
Debt to Income ratio is much slower. Again, in the loan amounts, we can clearly
see the effet of the 2008 GFC - the average loan size decreased significantly,
though the accepted debt-to-income ratio was not affected as much.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_5}
ggplot(data = subset(prosper_loans, LoanStatus != "Current"),
       aes(x = IncomeRange)) +
  geom_bar(position = "fill", aes(fill = LoanStatus)) +
  theme(axis.text.x = element_text(angle = -45, hjust = 0))

ggplot(data = subset(prosper_loans, LoanStatus != "Current"),
       aes(x = ProsperRating..Alpha.)) +
  geom_bar(position = "fill", aes(fill = LoanStatus))
```

There appears to be a larger proportion of Defaulted loans for customers with no
displayed income range, or an income of $0, but for any positive income, the
proportions seem about the same, with a slight but apparent downward trend in
the proportion of cancelled loans.

When comparing the statuses with the Prosper rating, a trend was more apparent.
As expected, AA ratings have the lowest proportion of charged off loans and
those that are past due, and the proportion of charged off loans steadily
increases as you look down the levels of Prosper rating.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I could see the loan amounts varying by ListingCategory and EmploymentStatus in
expected ways - certain loans such as Personal Loans, Cosmetic Procedures, and
Student Use loans show a lower median amount than other categories, whereas Debt
Consolidation shows the higher median. EmploymentStatus shows the natural result
of full-time employed borrowers receiving the largest loans

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

In this section, the relationship I found most interesting was the change in the
Debt to Income Ratio over time. Clearly inflation is going to influence the
absolute loan amounts, as see in the plot of loan amounts over time, but the
Debt to Income Ratio shows that in proportion to their income, borroweres are
borrowing even more. I especially expected this to decrease around the 2008 GFC,
but there is nothing to be see on the chart.

### What was the strongest relationship you found?

The relationship between the Prosper Rating and the proportion of completed
loans seems to be the strongest relationship that I can see. As the chart shows,
all of the "bad" statuses - e.g., Chargedoff, Defaulted, and Past Due - decrease
in proportion as the Prosper rating increases, and so with that the proportion
of Completed loans increases very clearly. Thankfully, this is one that I
expected.

# Multivariate Plots Section

Here I want to start exploring a few trends grouped by certain categorical
features, though my choices are limited as many features, such as Loan type and
Occupation, have far too many values to use colour as a differentiator.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
head(prosper_loans[,c(4,13,15,16,17,18,19,20,22,29,36,47,48)])
#ggpairs(prosper_loans[,c(4,13,15,16,17,20,22,29,36,47,48)])

ggplot(data = prosper_loans, aes(x = as.yearmon(LoanOriginationDate),
                                 y = DebtToIncomeRatio)) +
  geom_line(stat = "summary", fun.y = mean, aes(color = IsBorrowerHomeowner)) +
  scale_x_yearmon() +
  geom_smooth(aes(color = IsBorrowerHomeowner)) +
  #facet_wrap(~IsBorrowerHomeowner, ncol = 1) +
  ylim(0, .3)
```

Seems to be the same for both Yes and No. Not much of interest here.

Earlier I commented on the loans with low rates and poor credit ratings. Here I
would like to explore the types of loans that fall into these buckets.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_1a}
ggplot(data = subset(prosper_loans, 0 < prosper_loans$CreditScoreRangeLower &
                       prosper_loans$CreditScoreRangeLower <= 500),
       aes(x = CreditScoreRangeLowerFactor, y = BorrowerRate)) +
  # geom_jitter(aes(color = ListingCategory), alpha = 0.5, height = 0.5,
  #             width = 6)
  geom_boxplot() +
  facet_wrap(~ListingCategory, ncol = 1)

ggplot(data = subset(prosper_loans, 0 < prosper_loans$CreditScoreRangeLower &
                       prosper_loans$CreditScoreRangeLower <= 500),
       aes(x = CreditScoreRangeLowerFactor, y = BorrowerRate)) +
  # geom_jitter(aes(color = EmploymentStatus), alpha = 0.5, height = 0.5,
  #             width = 6)
  geom_boxplot() +
  facet_wrap(~EmploymentStatus, ncol = 1)
```

Unfortunately, they are all ListingCategory 0, which is "Not Available", and the
 employment status is not available, either. Let me try to extend the range a 
 little
 
```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_2}
ggplot(data = subset(prosper_loans, 500 < prosper_loans$CreditScoreRangeLower &
                       prosper_loans$CreditScoreRangeLower <= 600),
       aes(x = CreditScoreRangeLower, y = BorrowerRate)) +
  geom_jitter(aes(color = EmploymentStatus), alpha = 0.5, height = 0.5,
              width = 8)

ggplot(data = subset(prosper_loans, 500 < prosper_loans$CreditScoreRangeLower &
                       prosper_loans$CreditScoreRangeLower <= 600),
       aes(x = CreditScoreRangeLower, y = BorrowerRate)) +
  geom_jitter(aes(color = ListingCategory), alpha = 0.5, height = 0.5,
              width = 8)
```

These charts helps a little, though there are rather too many colours for my
taste. Perhaps a box plot grid will work better?

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_3}
ggplot(data = subset(prosper_loans, 500 < prosper_loans$CreditScoreRangeLower &
                       prosper_loans$CreditScoreRangeLower <= 600),
       aes(x = CreditScoreRangeLowerFactor, y = BorrowerRate)) +
  # geom_jitter(alpha = 0.5, height = 0.5, width = 8) +
  geom_boxplot() +
  facet_wrap(~EmploymentStatus)

ggplot(data = subset(prosper_loans, 500 < prosper_loans$CreditScoreRangeLower &
                       prosper_loans$CreditScoreRangeLower <= 600),
       aes(x = CreditScoreRangeLowerFactor, y = BorrowerRate)) +
  # geom_jitter(alpha = 0.5, height = 0.5, width = 8) +
  geom_boxplot() +
  facet_wrap(~ListingCategory)
```

Looking at employment status, what we can see now is that the low credit rating
and low borrower rate loans are not weighted towards one type in particular -
all I can say is that I am surprised that there are borrowers that are not
employed with a CreditScoreRangeLower of 520 and who received a rate of 0.1.

Looking at the loan type (ListingCategory) and comparing it to the histogram of
ListingCategory, what is interesting is the number of low credit score and low
rate loans for categories 4, 5, and 6, which have a very small number of loans
in total. These categories correspond to "Personal Loan", "Student Use", and
"Auto", respectively. I can understand why "Student Use" might allow for low
rates, as the borrowers may not have had time to develop a good credit score,
but I cannot answer the question as to why personal and auto loans are being
given with low rates to those with low credit scores.

How about the number of loans given for various credit ratings?
```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_4}
ggplot(data = prosper_loans, aes(x = as.yearmon(LoanOriginationDate))) +
  geom_area(stat = "count") +
  facet_wrap(~ProsperRating..Alpha.) +
  scale_x_yearmon() + 
  theme(axis.text.x = element_text(angle = -45, hjust = 0))

ggplot(data = prosper_loans, aes(x = as.yearmon(LoanOriginationDate))) +
  geom_line(stat = "count", aes(color = ProsperRating..Alpha.)) +
  scale_x_yearmon() + 
  theme(axis.text.x = element_text(angle = -45, hjust = 0))

ggplot(data = subset(prosper_loans,
                     prosper_loans$ProsperRating..Alpha. != "NA"),
       aes(x = as.yearmon(LoanOriginationDate))) +
  geom_line(stat = "count", aes(color = ProsperRating..Alpha.)) +
  scale_x_yearmon() + 
  theme(axis.text.x = element_text(angle = -45, hjust = 0))

ggplot(data = subset(prosper_loans,
                     prosper_loans$ProsperRating..Alpha. != "NA"),
       aes(x = as.yearmon(LoanOriginationDate))) +
  geom_line(stat = "count", aes(color = ProsperRating..Alpha.), size = 1.5) +
  scale_x_yearmon(limits = c(2013,2014)) +
  theme(axis.text.x = element_text(angle = -45, hjust = 0))
```

Straight away I can see that for all loans prior to roughly July 2009, there was
no Prosper rating, so I've focussed on the non-NA ones, and then zoomed in to
2013 and 2014. What I can see is that most Prosper ratings have seen an increase
during this period, except for the lowest ratings (HR, E, and D) which is
probably a positive trend (at least for the lenders), though of course it means
that it may have become harder for clients with lower credit scores to secure
loans.

Next I wanted to look at the average loan amount over time according to the
credit rating or Prosper rating of the borrower. Since it forms quite a jagged
line, I've also included a smoother version

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_5}
ggplot(data = subset(prosper_loans, CreditGrade != "NA"),
       aes(x = as.yearmon(LoanOriginationDate), y = LoanOriginalAmount)) +
  geom_line(stat = "summary", fun.y = mean, aes(color = CreditGrade)) +
  scale_x_yearmon() + 
  theme(axis.text.x = element_text(angle = -45, hjust = 0))

ggplot(data = subset(prosper_loans, CreditGrade != "NA"),
       aes(x = as.yearmon(LoanOriginationDate), y = LoanOriginalAmount)) +
  geom_smooth(aes(color = CreditGrade)) +
  scale_x_yearmon() + 
  theme(axis.text.x = element_text(angle = -45, hjust = 0))

ggplot(data = subset(prosper_loans, ProsperRating..Alpha. != "NA" ),
       aes(x = as.yearmon(LoanOriginationDate), y = LoanOriginalAmount)) +
  geom_line(stat = "summary", fun.y = mean,
            aes(color = ProsperRating..Alpha.)) +
  scale_x_yearmon() + 
  theme(axis.text.x = element_text(angle = -45, hjust = 0))

ggplot(data = subset(prosper_loans, ProsperRating..Alpha. != "NA" ),
       aes(x = as.yearmon(LoanOriginationDate), y = LoanOriginalAmount)) +
  geom_smooth(aes(color = ProsperRating..Alpha.)) +
  scale_x_yearmon() + 
  theme(axis.text.x = element_text(angle = -45, hjust = 0))
```

No one rating seems to follow a different pattern. In general, the higher the
rating, the larger the average loan amount, which is expected. The averages go
roughly up and down together.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the
investigation. Were there features that strengthened each other in terms of
looking at your feature(s) of interest?

There is nothing that I could see to strengthen the relationship between any of
the features of the interest. Splitting the changes in loan amount over time by
creditor grade, as above for example, did not highlight any real differences.
The only real possibility is the number of loans over time for the different
Prosper ratings - there may be a small difference between in the final months
of the data.

### Were there any interesting or surprising interactions between features?

I'll be honest, I can't see any real correlations in this data that are
interesting. All I've been able to find is that a better credit/Prosper rating
leads to a higher average loan amount, as they are more likely to pay it back,
and a lower rate. When I tried to look at the loans with a low rate but a bad
rating, I couldn't see any patterns in the employment status, and just a few
intriguing data points when looking at loan category.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
#This function will be used to find the mode of a few features
#Source: https://www.tutorialspoint.com/r/r_mean_median_mode.htm
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

ggplot(data = prosper_loans, aes(x = IncomeRange) ) +
  geom_bar(stat = "count") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  xlab("Income Range") +
  scale_y_continuous(labels=comma) +
  ylab("Number of Loans") +
  ggtitle("Number of Loans by Income Range")
```

### Description One
I chose this chart as I believe the quantum of these peer-to-peer loans is the
most interesting individual feature of the data, and a simple histogram shows it
best in one dimension. This chart show an expected peak around the average
income buckets ($25,000 - $74,999), but what is interesting to me is the
non-zero number of loans for borrowers unemployed or earning $0 - the lenders
must be quite confident in the borrowers' ability to repay.

The mean income for all of these points is
`r dollar_format()(mean(prosper_loans$StatedMonthlyIncome * 12))` which is based off the
StatedMonthlyIncome. The median is
`r dollar_format()(median(prosper_loans$StatedMonthlyIncome * 12))`, and the mode of the income
range is `r getmode(prosper_loans$IncomeRange)`.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(data = subset(prosper_loans, LoanStatus != "Current"),
       aes(x = ProsperRating..Alpha.)) +
  geom_bar(position = "fill", aes(fill = LoanStatus)) +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  xlab("Prosper Rating") +
  scale_y_continuous(labels=percent) +
  ylab("Percentage of Loans") +
  ggtitle("Loan Statuses by Prosper Rating")
```

### Description Two
I chose this chart as it shows the Prosper rating does seem to work well in
predicting borrowers who will pay off their loan, as demonstrated by a lower
proportion of charged off loans for the higher ratings. It is interesting to
look at the NA category, which is all loans originating in or before July 2009.
The percentage of charged off and defaulyed loans is quite high, and none are
past due because they are too old - they are either paid off or written off by
the time this data set was collated.

The most common Prosper Rating, after filering out the NAs,  is `r getmode(
subset(prosper_loans, !is.na(prosper_loans$ProsperRating..Alpha.)
)$ProsperRating..Alpha.)`.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(data = subset(prosper_loans,
                     prosper_loans$ProsperRating..Alpha. != "NA"),
       aes(x = as.yearmon(LoanOriginationDate))) +
  geom_line(stat = "count", aes(color = ProsperRating..Alpha.), size = 1.5) +
  scale_x_yearmon(limits = c(2013,2014)) +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  xlab("Loan Origination Date") +
  ylab("Number of Loans") +
  ggtitle("Number of Loans over Time by Prosper Rating")
```

### Description Three
I chose this chart to show the change in loan portfolio over time, according to
the distribution of Proser ratings. The different trends for different ratings
are clear to see, and it's interesting to see in this plot the increase in loan
numbers for the higher rated borrowers (B, A, and AA) in the later months
(August 2013 onwards), but the number of loans for lower rated borrowers stayed
far more steady. It's not a good trend if you have a low rating.

# Reflection

There are a very large number of variables in this set mostly on the number and
dollar amount of loans split by various categories. I expected clear changes
over time and across categories such as loan type (ListingCategory) and
employment status (EmploymentStatus), and whilst these changes are discernable,
there was nothing much surprising.

There were only two results that surprised me: In the chart of BorrowerRate and
CreditScoreRangeUpper/CreditScoreRangeLower, I expected a steady trend, but the
smoother showed two peaks around 0.06 and 0.3 for BorrowerRate, meaning that at
some decision point having a higher credit score gives you a higher rate.
Secondly, the smoothed line of LoanOriginationDate and DebtToIncomeRatio did not
drop after the 2008 GFC, but stayed fairly steady for a long period of time.

The first two of my three final charts show expected distributions - loan
numbers clustered around median income ranges, and loan statuses showing that
better Prosper ratings lead to fewer charged off loans. The third chart is a
little surprising, showing the differing trends between Prosper ratings in the 
later months of the data. I wonder if this is due to deliberate action by
Prosper?

One direction for future analysis could be to look at employment status, loan
type, and prosper rating all together over time to see if there is any
discernable causation - are the changes in ratings seen above due to a different
demographic (according to employment) taking out these loans, or is Prosper
perhaps becoming more stringent in their requirements?